---
title: "Module 6: Linear Regression — Fitting a Line"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: >
  When your predictor is a measured number rather than a group, you fit a line.
  Build a linear model with lm(), read the slope, intercept, R-squared, and
  p-value out of the summary, and say what each one means biologically.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(httr)
library(shiny)
library(BisonExplorR)   # coach_ui() / coach_server()

gradethis_setup()
tutorial_options(
  exercise.checker = gradethis::grade_learnr,
  exercise.completion = TRUE
)
knitr::opts_chunk$set(echo = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Resolve the student ID no matter which session scope a handler receives.
# VERIFIED 2026-07-25 (learnr 0.11.6): question() blocks are Shiny MODULES, so
# a question_submission handler gets a NAMESPACED session whose input$sid is
# EMPTY. The ID lives on the root session. This fails silently -- rows land in
# the sheet looking well-formed but unattributable, grade_module() drops them,
# and the student loses that score with no warning anywhere. Do not "simplify"
# this back to session$input$sid.
bx_get_sid <- function(session) {
  usable <- function(x) !is.null(x) && length(x) > 0 && nzchar(as.character(x)[1])
  sid <- tryCatch(session$rootScope()$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$userData$bx_sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  NULL   # log_attempt()'s %||% turns this into "unknown"
}

# Is this the GRADED deployment or a practice copy?
# Set BISONEXPLORR_GRADED=true in the deployed app's .Renviron (or set the
# option in its startup). Practice copies launched with run_tutorial() never
# see it, so they default to FALSE.
#
# WHY THE PRACTICE BRANCH SHOWS A VISIBLE BANNER RATHER THAN NOTHING: if this
# flag is ever missing on a deployed graded app, the ID box disappears and
# nothing logs -- every student silently scores zero. A banner makes a
# misconfigured deploy announce itself on the first screen instead of three
# weeks later in the response sheet.
bx_is_graded <- function() {
  isTRUE(getOption(
    "BisonExplorR.graded",
    identical(tolower(Sys.getenv("BISONEXPLORR_GRADED")), "true")
  ))
}

log_attempt <- function(student_id, exercise_id, correct) {
  # Practice copies do not log. Without this, practice attempts POST with
  # student_id = "unknown" and fill the response sheet with orphan rows
  # that look like a bug in grade_module().
  if (!bx_is_graded()) return(invisible(NULL))

  tryCatch({
    httr::POST(
      url = "https://docs.google.com/forms/d/e/1FAIpQLSfmm3cUcoaWWYBb0zk7rbowSpgUL13nQc-Dxn7K8zQ30GdykA/formResponse",
      body = list(
        "entry.1413675438" = as.character(Sys.time()),
        "entry.1257984687" = as.character(student_id %||% "unknown"),
        "entry.1315894547" = "module6",
        "entry.1172625272" = as.character(exercise_id %||% "unknown"),
        "entry.1835356043" = as.character(correct %||% "NA")
      ),
      encode = "form"
    )
  }, error = function(e) message("Logging failed: ", e$message))
}

# Package dataset — always available to every exercise
algae <- read.csv(system.file("extdata", "algae_light.csv", package = "BisonExplorR"))

# The fitted model, pre-made so later exercises can use it (learnr sandboxes
# each exercise, so anything shared has to be built here).
fit <- lm(co2_drawdown_ppm ~ light_umol, data = algae)
```

```{r event-setup, context="server-start"}
event_register_handler("exercise_result", function(session, event, data) {
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = data$feedback$correct
  )
})
```

```{r coach-server, context="server"}
# One coach_server() per coached exercise. The exercise_id string must match a
# key in the server-side exercise_registry.R ("moduleN-qM", hyphenated).
# get_tutorial_state() is what feeds the coach the student's own submission.
coach_server("coach_q2", shiny::reactive(learnr::get_tutorial_state("q2")), "module6-q2")
coach_server("coach_q4", shiny::reactive(learnr::get_tutorial_state("q4")), "module6-q4")
coach_server("coach_q5", shiny::reactive(learnr::get_tutorial_state("q5")), "module6-q5")
```

## Before you start

```{r student-id, echo=FALSE}
uiOutput("sid_ui")
```

```{r sid-server, context="server"}
# The ID box exists ONLY in the graded deployment. See bx_is_graded() in setup.
output$sid_ui <- renderUI({
  if (!bx_is_graded()) {
    return(shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #C05621; background:#FDF6EC;"),
      shiny::strong("Practice mode."),
      " Nothing here is recorded and this copy does not count for credit.",
      " To do this module for credit, run ",
      shiny::tags$code('launch_graded_tutorial("module6")'),
      " instead."
    ))
  }
  shiny::tagList(
    shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #1F3864; background:#F2F4F7;"),
      shiny::strong("Graded copy."),
      " Your work on this module is recorded, so enter your student ID before",
      " you begin. Use the same ID every time \u2014 a different ID reads as a",
      " different student and your work will not be combined."
    ),
    shiny::textInput("sid",
      "Student ID (the same one you used in Module 1):", value = ""),
    shiny::div(id = "sid_warn",
      style = "color:#C05621; font-weight:600; margin:-6px 0 8px 0;",
      "Enter your ID to continue."),
    # Client-side gate on the section's Continue button. This prevents the
    # ACCIDENTAL case (a student who forgets), not a determined one -- anyone
    # with devtools can re-enable the button. The real protection against
    # unattributed rows is that grade_module() drops blank IDs.
    shiny::tags$script(shiny::HTML("
      (function () {
        function apply() {
          var sid = document.getElementById('sid');
          if (!sid) return;
          var sec = sid.closest('.section');
          if (!sec) return;
          var btns = sec.querySelectorAll('.topicActions button, .continueButton');
          var ok = sid.value.trim().length > 0;
          btns.forEach(function (b) {
            b.disabled = !ok;
            b.style.opacity = ok ? '' : '0.45';
            b.style.cursor  = ok ? '' : 'not-allowed';
          });
          var w = document.getElementById('sid_warn');
          if (w) w.style.display = ok ? 'none' : 'block';
        }
        document.addEventListener('input', function (e) {
          if (e.target && e.target.id === 'sid') apply();
        });
        // learnr renders sections progressively, so the Continue button may not
        // exist yet when this script first runs.
        new MutationObserver(apply)
          .observe(document.body, { childList: true, subtree: true });
        apply();
      })();
    "))
  )
})

# Cache the ID on userData as a second, independent path to it. bx_get_sid()
# checks rootScope() first and falls back to this.
shiny::observeEvent(input$sid, {
  if (!is.null(input$sid) && nzchar(input$sid)) {
    session$userData$bx_sid <- input$sid
  }
}, ignoreInit = TRUE)
```


In Module 5 you compared **groups**. But a lot of biology isn't about groups at all — it's about
one measured number predicting another. Does more light mean more photosynthesis? Does a bigger
animal have a slower heart? Those questions need a **line**, not a group comparison.

We'll work with `algae`, data from an algae-bead experiment much like the one you ran: 36 vials of
beads, each held at a different **light intensity**, with the **CO~2~ drawn down** out of the
indicator solution measured after 30 minutes.

```{r peek, exercise=TRUE}
head(algae)
```

| Column | What it is | Units |
|---|---|---|
| `vial_id` | which vial | — |
| `light_umol` | light intensity the vial was held at | µmol photons m^-2^ s^-1^ |
| `co2_drawdown_ppm` | CO~2~ removed from the solution in 30 min | ppm |

A **positive** drawdown means the beads pulled CO~2~ out of the solution — net photosynthesis.
A **negative** drawdown means CO~2~ went *up* — the beads released more by respiring than they
took up. Hold onto that; it matters later.

By the end of this module you should be able to:

- Recognize when a question needs a regression rather than a group comparison
- Fit a linear model with `lm()` and read its `summary()`
- Say what the **slope**, the **intercept**, the **R^2^**, and the **p-value** each mean
- Report a regression result the way a biologist would

---

## The map: which test, when

Keep this table in view — it is the same in every statistics module. The last column tells you
where each test is taught.

| Biological question | Predictor(s) | Test | R function | Module |
|---|---|---|---|---|
| Do **two** groups differ? | one factor, 2 levels | t-test | `t.test()` | 5 |
| Do **three or more** groups differ? | one factor, 3+ levels | one-way ANOVA | `aov()` + `summary()` | 5 |
| Does a **continuous** variable predict the response? | one continuous | linear regression | `lm()` + `summary()` | **6** |
| Do groups differ across **two factors**? | two factors | two-way ANOVA | `aov()` + `summary()` | 7 |
| Do groups differ **after controlling for** a continuous variable? | factor + covariate | ANCOVA | `aov()` + `summary()` | 8 |

**Where you are:** Module 6 — one continuous predictor.

### Pick the test

```{r pick_test, echo=FALSE}
question(
  "You want to know whether light intensity predicts CO~2~ drawdown. Light was set to nine different values from 0 to 400 µmol. Which test?",
  answer("Linear regression", correct = TRUE, message = "Right. Light is a measured number, not a label — so you want a slope, not a group difference."),
  answer("One-way ANOVA with nine groups", message = "You could force it, but you would throw away the ordering. ANOVA treats 50 and 400 as two unrelated labels; regression uses the fact that 400 is eight times 50."),
  answer("t-test", message = "A t-test compares two groups. There are no groups here — light is continuous."),
  allow_retry = TRUE
)
```

---

## You have already drawn this line

In Module 4 you added a trend line to a scatter plot with `geom_smooth(method = "lm")`. That line
was already a regression — `ggplot` fitted it quietly and drew it. This module opens the hood:
the same line, but now you get the **numbers** behind it.

Always start by looking. A regression on data you haven't plotted is a regression you can't trust.

### Your turn

Build a scatter plot of `co2_drawdown_ppm` (y) against `light_umol` (x), with `geom_point()`.

```{r q1, exercise=TRUE}
ggplot(algae, aes(x = ___, y = ___)) +
  ___
```

```{r q1-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That didn't return a plot. Start with ggplot(algae, aes(...)) and add geom_point().")
  }
  mapped <- names(.result$mapping)
  xvar <- rlang::as_label(.result$mapping$x %||% quote(NA))
  yvar <- rlang::as_label(.result$mapping$y %||% quote(NA))
  geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  if (!"GeomPoint" %in% geoms) {
    fail("You need a points layer — add geom_point().")
  }
  if (!grepl("light_umol", xvar) || !grepl("co2_drawdown_ppm", yvar)) {
    fail("Check your axes: light_umol is the predictor (x), co2_drawdown_ppm is the response (y).")
  }
  pass("There's the relationship — it climbs, roughly in a straight line, and the lowest-light vials sit below zero.")
})
```

Notice two things in that plot. The cloud **rises** — more light, more drawdown. And at the
left-hand edge, several points sit **below zero**. Those are the dark vials, releasing CO~2~.

---

## Fitting the model: `lm()` {data-allow-skip="true"}

A linear model is one function call. The formula reads **response `~` predictor** — the same
`y ~ x` shape you used for `t.test()` in Module 5:

```r
lm(response ~ predictor, data = your_data)
```

On its own, `lm()` prints almost nothing useful. Wrap it in `summary()` — that's where the whole
readout lives. This is the house pattern for every model in this course: **`lm()` then
`summary()`**.

### Your turn

Fit a model predicting `co2_drawdown_ppm` from `light_umol` using the `algae` data, and pass it
straight to `summary()`.

```{r q2, exercise=TRUE}
summary(lm(___ ~ ___, data = ___))
```

```{r q2-check}
grade_this({
  if (!inherits(.result, "summary.lm")) {
    fail("That didn't return a model summary. Wrap your lm() call in summary().")
  }
  cf <- coef(.result)
  if (!"light_umol" %in% rownames(cf)) {
    fail("Check your formula — it should be co2_drawdown_ppm ~ light_umol (response on the left).")
  }
  if (abs(cf["light_umol", "Estimate"] - 0.12175) > 0.001) {
    fail("The formula looks reversed. Response goes on the left of the ~, predictor on the right.")
  }
  pass("That readout is the whole model. Now let's take it apart line by line.")
})
```

Your output should look like this:

```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)
(Intercept) -14.466667   2.087800  -6.929 5.50e-08 ***
light_umol    0.121750   0.008770  13.882 1.44e-15 ***

Residual standard error: 6.794 on 34 degrees of freedom
Multiple R-squared:  0.85,	Adjusted R-squared:  0.8456
F-statistic: 192.7 on 1 and 34 DF,  p-value: 1.441e-15
```

Four numbers matter. We'll take them one at a time.

```{r coach-ui-q2, echo=FALSE}
coach_ui("coach_q2", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.


---

## The slope

The slope is the row named after your predictor — here `light_umol`, estimate **0.1218**.

A slope always carries **units**: *response units per one unit of predictor*. So:

> For every extra 1 µmol m^-2^ s^-1^ of light, CO~2~ drawdown increases by about **0.12 ppm**.

Per-unit slopes are often awkwardly small. Scale them to something a reader can picture — over a
100 µmol increase in light, drawdown rises by about **12 ppm**.

```{r slope_units, echo=FALSE}
question(
  "The slope is 0.1218. What are its units?",
  answer("ppm per µmol m^-2^ s^-1^", correct = TRUE, message = "Yes — response units over predictor units. Always state both."),
  answer("ppm", message = "Not quite. A slope is a *rate*: how much y changes per unit of x. It needs both units."),
  answer("µmol m^-2^ s^-1^", message = "That's the predictor's units alone. A slope combines both."),
  answer("It has no units — it's a correlation", message = "That's r or R^2^, which are unitless. A regression slope is not."),
  allow_retry = TRUE
)
```

### Your turn

`coef()` pulls the two coefficients out as a named numeric vector — handy when you want the
numbers themselves rather than the printed table. The fitted model is already available as `fit`.

```{r q3, exercise=TRUE}
# Return the coefficients of the fitted model.
___(fit)
```

```{r q3-check}
grade_this({
  if (!is.numeric(.result) || length(.result) != 2) {
    fail("Expected two numbers — the intercept and the slope. Try coef(fit).")
  }
  if (!all(c("(Intercept)", "light_umol") %in% names(.result))) {
    fail("Use coef(fit) — it returns both coefficients with their names.")
  }
  pass("Intercept -14.47, slope 0.1218. The same two numbers from the summary table, now as a vector you could do arithmetic on.")
})
```

---

## The intercept: what happens in the dark

The intercept is the model's prediction when the predictor equals **zero** — here, drawdown at
zero light. It comes out at **-14.47 ppm**, and it is significantly different from zero
(p = 5.5 × 10^-8^).

That negative value is not an error. It is the biology:

> In darkness the beads cannot photosynthesise, but they are still **respiring** — so CO~2~ in
> the vial *rises* rather than falls. The intercept is an estimate of the algae's dark respiration
> rate.

This is worth dwelling on, because intercepts are often meaningless. If you regressed heart rate
on body mass, the intercept would be the predicted heart rate of an animal weighing zero
kilograms — a number with no biological referent at all. **Always ask whether zero is a real,
reachable value for your predictor before you interpret an intercept.** Here it is: a vial in the
dark is a perfectly ordinary vial.

```{r intercept_meaning, echo=FALSE}
question(
  "A classmate regresses leaf area on plant height across 40 seedlings and reports: 'The intercept is 3.1 cm^2^, so a plant of zero height has 3.1 cm^2^ of leaf.' What's wrong?",
  answer("A plant of zero height doesn't exist, so the intercept has no biological meaning here — it's just where the fitted line happens to cross the axis.", correct = TRUE, message = "Exactly. The intercept is always mathematically defined; whether it *means* anything depends on whether zero is reachable."),
  answer("Nothing — that's the correct reading of an intercept.", message = "The arithmetic is right but the biology isn't. Zero height is not an achievable plant."),
  answer("The intercept should always be zero.", message = "No — there's no rule forcing an intercept through the origin, and forcing one usually distorts the fit."),
  answer("They should have used the slope instead.", message = "The slope is a different quantity. The issue is specifically whether zero is a meaningful predictor value."),
  allow_retry = TRUE
)
```

---

## R^2^: how much of the variation is explained {data-allow-skip="true"}

**R^2^** answers a different question from the slope. The slope says *which way and how steeply*;
R^2^ says *how tightly the points hug the line*.

It runs from 0 to 1, and reads as a proportion of variation explained. Here R^2^ = **0.85**:

> About **85%** of the variation in CO~2~ drawdown is explained by light intensity. The other 15%
> is everything else — differences between beads, pipetting error, temperature drift.

A high R^2^ does **not** mean the relationship is important, and a low one doesn't mean it's
absent. A slope can be tiny and precise (high R^2^, trivial effect) or large and noisy (low R^2^,
big effect). Report both.

### Your turn

`summary(fit)` returns an object you can reach into with `$`. The element you want is
`r.squared`.

```{r q4, exercise=TRUE}
# Pull just the R-squared value out of the model summary.
summary(fit)$___
```

```{r q4-check}
grade_this({
  if (!is.numeric(.result) || length(.result) != 1) {
    fail("Expected a single number. Try summary(fit)$r.squared.")
  }
  if (abs(.result - 0.85) > 0.005) {
    fail("Close, but that isn't R-squared. Note that $adj.r.squared is a different (slightly smaller) value.")
  }
  pass("0.85 — light intensity accounts for about 85% of the variation in drawdown.")
})
```

```{r coach-ui-q4, echo=FALSE}
coach_ui("coach_q4", "Stuck? Ask the Coach")
```

> Ask the Coach if you are unsure *which* R-squared the exercise wants.


---

## The p-value: is the slope real?

The p-value on the slope row tests one specific claim: **the true slope is zero** — that is, that
light tells you nothing about drawdown. Here it is 1.44 × 10^-15^, far below 0.05, so we reject
that claim.

Note what this p-value is *not* about. It doesn't say the relationship is strong (that's R^2^), or
that a line is the right shape (that's your scatter plot's job), or that light *causes* the
drawdown — though here it does, because light was **experimentally set** rather than merely
observed. Causation comes from the design, never from the model.

```{r pval_meaning, echo=FALSE}
question(
  "The slope's p-value is 1.44e-15. Which statement is correct?",
  answer("There is very strong evidence that the true slope is not zero.", correct = TRUE, message = "Right — that's precisely the claim being tested."),
  answer("There is a 1.44e-15 probability that the results are due to chance.", message = "A common misreading. The p-value is the probability of data this extreme *if* the slope were truly zero — not the probability that chance is the explanation."),
  answer("The relationship is very strong.", message = "Strength is R^2^ (0.85 here). A p-value can be tiny for a weak relationship if the sample is large."),
  answer("85% of the variation is explained.", message = "That's the R^2^, a different number in the same output."),
  allow_retry = TRUE
)
```

---

## Putting it back on the figure {data-allow-skip="true"}

Now close the loop. In Module 4 you annotated a trend line with `stat_regline_equation()` and
`stat_cor()` from **ggpubr** — those two functions are drawing exactly the numbers you just
extracted by hand.

### Your turn

Rebuild the scatter plot, add a straight-line fit with `geom_smooth(method = "lm")`, and annotate
it with `stat_regline_equation()`.

```{r q5, exercise=TRUE}
ggplot(algae, aes(x = light_umol, y = co2_drawdown_ppm)) +
  geom_point() +
  ___(method = "lm") +
  ___() +
  labs(x = "Light intensity (µmol m-2 s-1)",
       y = "CO2 drawdown (ppm)")
```

```{r q5-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That didn't return a plot.")
  }
  geoms <- sapply(.result$layers, function(l) class(l$geom)[1])
  if (!"GeomPoint" %in% geoms) {
    fail("Keep the points layer — geom_point().")
  }
  if (!any(c("GeomSmooth", "GeomLine") %in% geoms)) {
    fail("Add the fitted line with geom_smooth(method = \"lm\").")
  }
  if (!any(grepl("Text|Label", geoms))) {
    fail("Add stat_regline_equation() so the equation is printed on the figure.")
  }
  pass("The equation on that figure is y = -14 + 0.12x — your slope and intercept, drawn.")
})
```

```{r coach-ui-q5, echo=FALSE}
coach_ui("coach_q5", "Stuck? Ask the Coach")
```

> Ask the Coach if a layer is missing and you cannot tell which one.


---

## Reporting a regression

A regression result in a biology paper needs four things: the **direction and size** of the slope
with units, the **R^2^**, the **test statistic and p-value**, and the **sample size**. Something
like:

> CO~2~ drawdown increased significantly with light intensity, by 0.12 ppm per µmol m^-2^ s^-1^
> (linear regression: R^2^ = 0.85, F~1,34~ = 192.7, p < 0.001, n = 36). The negative intercept
> (-14.5 ppm) indicates net CO~2~ release under darkness, consistent with continued respiration.

Notice that the sentence leads with the **biology** and puts the statistics in parentheses. The
numbers support the claim; they aren't the claim.

```{r reporting_check, echo=FALSE}
question(
  "Which of these is missing from the report sentence above, if you were writing it for a lab report?",
  answer("Nothing — direction, size, units, R^2^, test statistic, p, and n are all there.", correct = TRUE, message = "Correct. Use it as your template."),
  answer("The mean drawdown", message = "A mean isn't part of a regression report — the line replaces it."),
  answer("The standard error of every point", message = "Not needed; the residual error is already summarised by R^2^ and the F-test."),
  allow_retry = TRUE
)
```

---

## You made it

In this module you learned to:

- tell a regression question from a group-comparison question
- fit a model with `lm(response ~ predictor, data = ...)` and read it with `summary()`
- interpret the **slope** with its units, and check whether the **intercept** is biologically meaningful
- read **R^2^** as variation explained, separately from the **p-value** on the slope
- extract coefficients with `coef()` and R^2^ with `summary(fit)$r.squared`
- annotate a fitted line on a figure, and report the result in a sentence

Next up is **Module 7**, which goes back to groups — but two grouping factors at once, and the
possibility that they interact. After that, **Module 8** brings the two halves together: ANCOVA,
which is this module's regression *plus* a grouping factor.
