---
title: "Module 4: Data Wrangling and Visualization"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: "Reshape a dataset with dplyr and build figures with ggplot2, including a fitted regression line."
---

```{css bx-style, echo=FALSE}
/* Sidebar is a progress display, not navigation.                   */
/* allow_skip: false gates the Continue button but NOT the sidebar, */
/* so without this a student can click any topic and land there     */
/* having completed nothing.                                        */
/* Disabling the <li>, not the <a>: the <nav> is a shiny-bound-input */
/* that delegates clicks, so an inert <a> just passes the click up.  */
.topicsList .nav-pills .topic { pointer-events: none; }
/* learnr paints a topicProgress.png sprite as the <li> background. */
/* With the sidebar inert it reads as clutter.                     */
.topicsList .nav-pills .topic { background-image: none; }
.topicsList .nav-pills .topic:not(.current) > .nav-link { opacity: 0.5; }

/* Start Over sits in .topicsFooter, outside .nav-pills, so it stays */
/* clickable. The Previous button is deliberately NOT hidden --      */
/* with the sidebar inert it is the only way back to review.         */
```


```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(BisonExplorR)   # coach_ui() / coach_server()
library(gradethis)
library(httr)
library(dplyr)
library(ggplot2)
library(ggpubr)

gradethis_setup()

tutorial_options(
  exercise.checker = gradethis::grade_learnr,
  exercise.completion = TRUE
)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Resolve the student ID no matter which session scope a handler receives.
# VERIFIED 2026-07-25 (learnr 0.11.6): question() blocks are Shiny MODULES, so
# a question_submission handler gets a NAMESPACED session whose input$sid is
# EMPTY. The ID lives on the root session. This fails silently -- rows land in
# the sheet looking well-formed but unattributable, grade_module() drops them,
# and the student loses that score with no warning anywhere. Do not "simplify"
# this back to session$input$sid.
bx_get_sid <- function(session) {
  usable <- function(x) !is.null(x) && length(x) > 0 && nzchar(as.character(x)[1])
  sid <- tryCatch(session$rootScope()$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$userData$bx_sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  NULL   # log_attempt()'s %||% turns this into "unknown"
}

# Is this the GRADED deployment or a practice copy?
# Set BISONEXPLORR_GRADED=true in the deployed app's .Renviron (or set the
# option in its startup). Practice copies launched with run_tutorial() never
# see it, so they default to FALSE.
#
# WHY THE PRACTICE BRANCH SHOWS A VISIBLE BANNER RATHER THAN NOTHING: if this
# flag is ever missing on a deployed graded app, the ID box disappears and
# nothing logs -- every student silently scores zero. A banner makes a
# misconfigured deploy announce itself on the first screen instead of three
# weeks later in the response sheet.
bx_is_graded <- function() {
  isTRUE(getOption(
    "BisonExplorR.graded",
    identical(tolower(Sys.getenv("BISONEXPLORR_GRADED")), "true")
  ))
}

log_attempt <- function(student_id, exercise_id, correct) {
  # Practice copies do not log. Without this, practice attempts POST with
  # student_id = "unknown" and fill the response sheet with orphan rows
  # that look like a bug in grade_module().
  if (!bx_is_graded()) return(invisible(NULL))

  tryCatch({
    httr::POST(
      url = "https://docs.google.com/forms/d/e/1FAIpQLSfmm3cUcoaWWYBb0zk7rbowSpgUL13nQc-Dxn7K8zQ30GdykA/formResponse",
      body = list(
        "entry.1413675438" = as.character(Sys.time()),
        "entry.1257984687" = as.character(student_id %||% "unknown"),
        "entry.1315894547" = "module4",   # tutorial_id for this module
        "entry.1172625272" = as.character(exercise_id %||% "unknown"),
        "entry.1835356043" = as.character(correct %||% "NA")
      ),
      encode = "form"
    )
  }, error = function(e) message("Logging failed: ", e$message))
}

# Package dataset — always available to every exercise
mammals <- read.csv(system.file("extdata", "mammals_data.csv", package = "BisonExplorR"))
```

```{r event-setup, context="server-start"}
event_register_handler("exercise_result", function(session, event, data) {
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = data$feedback$correct
  )
})
```

```{r coach-server, context="server"}
# One coach_server() per coached exercise. The third argument must match a
# key in the server-side exercise_registry.R ("moduleN-qM", HYPHENATED) --
# an unknown key does not error, it silently returns generic context.
# The Shiny id must match the coach_ui() call below it, or the card renders
# and never responds.
BisonExplorR::coach_server("coach_q5", shiny::reactive(learnr::get_tutorial_state("q5")), "module4-q5")
BisonExplorR::coach_server("coach_q10", shiny::reactive(learnr::get_tutorial_state("q10")), "module4-q10")
```

## Before you start

```{r student-id, echo=FALSE}
uiOutput("sid_ui")
```

```{r sid-server, context="server"}
# The ID box exists ONLY in the graded deployment. See bx_is_graded() in setup.
output$sid_ui <- renderUI({
  if (!bx_is_graded()) {
    return(shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #C05621; background:#FDF6EC;"),
      shiny::strong("Practice mode."),
      " Nothing here is recorded and this copy does not count for credit.",
      " To do this module for credit, run ",
      shiny::tags$code('launch_graded_tutorial("module4")'),
      " instead."
    ))
  }
  shiny::tagList(
    shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #1F3864; background:#F2F4F7;"),
      shiny::strong("Graded copy."),
      " Your work on this module is recorded, so enter your student ID before",
      " you begin. Use the same ID every time \u2014 a different ID reads as a",
      " different student and your work will not be combined."
    ),
    shiny::textInput("sid",
      "Student ID (the same one you used in Module 1):", value = ""),
    shiny::div(id = "sid_warn",
      style = "color:#C05621; font-weight:600; margin:-6px 0 8px 0;",
      "Enter your ID to continue."),
    # Client-side gate on the section's Continue button. This prevents the
    # ACCIDENTAL case (a student who forgets), not a determined one -- anyone
    # with devtools can re-enable the button. The real protection against
    # unattributed rows is that grade_module() drops blank IDs.
    shiny::tags$script(shiny::HTML("
      (function () {
        function apply() {
          var sid = document.getElementById('sid');
          if (!sid) return;
          var sec = sid.closest('.section');
          if (!sec) return;
          var btns = sec.querySelectorAll('.topicActions button, .continueButton');
          var ok = sid.value.trim().length > 0;
          btns.forEach(function (b) {
            b.disabled = !ok;
            b.style.opacity = ok ? '' : '0.45';
            b.style.cursor  = ok ? '' : 'not-allowed';
          });
          var w = document.getElementById('sid_warn');
          if (w) w.style.display = ok ? 'none' : 'block';
        }
        document.addEventListener('input', function (e) {
          if (e.target && e.target.id === 'sid') apply();
        });
        // learnr renders sections progressively, so the Continue button may not
        // exist yet when this script first runs.
        new MutationObserver(apply)
          .observe(document.body, { childList: true, subtree: true });
        apply();
      })();
    "))
  )
})

# Cache the ID on userData as a second, independent path to it. bx_get_sid()
# checks rootScope() first and falls back to this.
shiny::observeEvent(input$sid, {
  if (!is.null(input$sid) && nzchar(input$sid)) {
    session$userData$bx_sid <- input$sid
  }
}, ignoreInit = TRUE)
```

Work through each section in order — a **Continue** button appears once you answer correctly.

In Module 2 you imported a CSV of goldenrod data and learned to inspect it with `head()`, `str()`, and
`summary()`. This module picks up right there: now that you can *load* a dataset, you'll learn to
*reshape* it and *plot* it — the two skills you'll use in almost every 204 lab.

We'll work with `mammals`, a dataset of 29 species with their body mass, resting heart rate, and
taxonomic group. It's already loaded for you.

```{r peek, echo=TRUE}
head(mammals)
```

By the end of this module you should be able to:

- Chain steps together with the pipe, `%>%`
- Use `filter()`, `select()`, `mutate()`, `group_by()`, and `summarise()`
- Calculate a standard error and add SE bars to a figure
- Build scatter plots, boxplots, and bar charts with `ggplot2`

---

## The pipe: `%>%`

Almost every wrangling step in R can be written as: *take a dataset, do something to it.* The
**pipe** `%>%` lets you chain those steps left to right, like reading a recipe:

```
mammals %>%        # start with the mammals data
  filter(...) %>%   # then keep some rows
  summarise(...)    # then summarise what's left
```

Read `%>%` as the word **"then."** Whatever is on the left is handed to the function on the right
as its first argument. You'll use it in every exercise below.

---

## `filter()` — keep the rows you want

`filter()` keeps only the rows that meet a condition. Conditions use comparisons like `==` (equal
to), `>`, `<`, and `!=` (not equal to).

### Your turn

Keep only the rows where `group` is `"Rodent"`. (Note the double `==` and the quotes around text.)

```{r q1, exercise=TRUE}
# Keep only the rows where group is "Rodent".
mammals %>%
  filter()
```

```{r q1-check}
grade_this({
  correct <- mammals %>% filter(group == "Rodent")
  if (isTRUE(all.equal(.result, correct, check.attributes = FALSE))) {
    pass("Nice — eight rodent rows, everything else dropped.")
  }
  fail('Not quite. Inside filter(), write the condition group == "Rodent".')
})
```

---

## `select()` — keep the columns you want

Where `filter()` chooses **rows**, `select()` chooses **columns**. List the columns you want to
keep, separated by commas.

### Your turn

Keep just two columns: `species` and `heart_rate_bpm`, in that order.

```{r q2, exercise=TRUE}
# Keep only the species and heart_rate_bpm columns.
mammals %>%
  select()
```

```{r q2-check}
grade_this({
  correct <- mammals %>% select(species, heart_rate_bpm)
  if (isTRUE(all.equal(.result, correct, check.attributes = FALSE))) {
    pass("Two columns, in the order you asked for.")
  }
  fail("Not quite. Select species and heart_rate_bpm.")
})
```

---

## `mutate()` — add or change a column

`mutate()` creates a new column (or overwrites an existing one) from the columns you already have.
The rest of the dataset comes along unchanged.

### Your turn

Body mass is stored in kilograms. Add a new column called `body_mass_g` that expresses it in
**grams** (there are 1000 grams in a kilogram).

```{r q3, exercise=TRUE}
# Add a column body_mass_g equal to body_mass_kg * 1000.
mammals %>%
  mutate()
```

```{r q3-check}
grade_this({
  correct <- mammals %>% mutate(body_mass_g = body_mass_kg * 1000)
  if (isTRUE(all.equal(.result, correct, check.attributes = FALSE))) {
    pass("Grams now sit right alongside kilograms.")
  }
  fail("Not quite. Add body_mass_g = body_mass_kg * 1000 inside mutate().")
})
```

---

## `group_by()` + `summarise()` — collapse to group summaries

This pair is the workhorse of data analysis. `group_by()` splits the data into groups; `summarise()`
then computes one value per group. Together they turn 29 rows of mammals into one row per `group`.

### Your turn

Find the **mean heart rate for each group**. Call the summary column `mean_hr`.

```{r q4, exercise=TRUE}
# For each group, calculate the mean of heart_rate_bpm. Call it mean_hr.
mammals %>%
  group_by(group) %>%
  summarise()
```

```{r q4-check}
grade_this({
  correct <- mammals %>%
    group_by(group) %>%
    summarise(mean_hr = mean(heart_rate_bpm))
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct), check.attributes = FALSE))) {
    pass("Four groups, four means — rodents fastest, ungulates slowest.")
  }
  fail("Not quite. Inside summarise(), write mean_hr = mean(heart_rate_bpm).")
})
```

---

## Standard error — how precise is that mean? {data-allow-skip="true"}

A group mean is only an estimate. The **standard error (SE)** tells you how far that estimate is
likely to sit from the true mean. It's the standard deviation divided by the square root of the
sample size:

$$ SE = \frac{sd}{\sqrt{n}} $$

In R, use `sd(x)` for the standard deviation and `length(x)` for the sample size *n*. (Use
`length()`, not `n()` — this is the convention we use across all of 204.)

### Your turn

Keep the `mean_hr` column from before, and **add a second column** `se_hr` for the standard error
of `heart_rate_bpm`.

```{r q5, exercise=TRUE}
# Add a second summary column, se_hr, for the standard error of heart_rate_bpm.
# SE = sd(x) / sqrt(length(x)).  Use length(), not n().
mammals %>%
  group_by(group) %>%
  summarise(mean_hr = mean(heart_rate_bpm))
```

```{r q5-check}
grade_this({
  correct <- mammals %>%
    group_by(group) %>%
    summarise(mean_hr = mean(heart_rate_bpm),
              se_hr   = sd(heart_rate_bpm) / sqrt(length(heart_rate_bpm)))
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct), check.attributes = FALSE))) {
    pass("That se_hr column is exactly what your error bars will come from later.")
  }
  fail("Not quite. Add se_hr = sd(heart_rate_bpm) / sqrt(length(heart_rate_bpm)).")
})
```

```{r coach-ui-q5, echo=FALSE}
BisonExplorR::coach_ui("coach_q5", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

---

## Visualizing with `ggplot2` {data-allow-skip="true"}

Every `ggplot` is built from three pieces:

1. `ggplot(data, aes(x = ..., y = ...))` — the data and which columns map to which axes
2. a **geom** layer added with `+` — `geom_point()`, `geom_boxplot()`, `geom_col()`, ...
3. optional styling — axis labels with `labs()`, a clean look with `theme_classic()`

A quick note on the x-axis below: body mass runs from a 21-gram mouse to a 4-tonne elephant, so we
plot `log_mass` (the base-10 log of body mass, already in the dataset). Logging spreads those points
out instead of crushing them all against zero.

### Your turn: a scatter plot

Finish this scatter of `log_mass` (x) against `heart_rate_bpm` (y) by adding `geom_point()` with `+`.

```{r q6, exercise=TRUE}
# Add geom_point() to draw the points.
ggplot(mammals, aes(x = log_mass, y = heart_rate_bpm))
```

```{r q6-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That isn't a plot yet — start from the ggplot() line and add a geom with +.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomPoint" %in% geoms) {
    pass("A clear downward trend: bigger mammals beat slower. That's allometry.")
  }
  fail("Add + geom_point() to draw the points.")
})
```

### A boxplot by group

`geom_boxplot()` compares a numeric variable across categories. Put `group` on x and
`heart_rate_bpm` on y, then add the boxplot geom.

```{r q7, exercise=TRUE}
# Add geom_boxplot() to compare heart rate across groups.
ggplot(mammals, aes(x = group, y = heart_rate_bpm))
```

```{r q7-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("Start from the ggplot() line and add + geom_boxplot().")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomBoxplot" %in% geoms) {
    pass("One box per group — the spread of rodent heart rates dwarfs the others.")
  }
  fail("Add + geom_boxplot().")
})
```

### A bar chart with SE bars

This is the figure you'll build again and again in the Daphnia and cardiovascular labs: a bar for
each group mean, with error bars showing the standard error. The summary table is built for you —
add `geom_col()` for the bars, then `geom_errorbar()` for the SE.

```{r q8, exercise=TRUE}
# The summary table (mean and SE per group) is ready:
hr_summary <- mammals %>%
  group_by(group) %>%
  summarise(mean_hr = mean(heart_rate_bpm),
            se_hr   = sd(heart_rate_bpm) / sqrt(length(heart_rate_bpm)))

# Add geom_col() for the bars, then geom_errorbar() with
#   ymin = mean_hr - se_hr  and  ymax = mean_hr + se_hr.
ggplot(hr_summary, aes(x = group, y = mean_hr)) +
  labs(x = "Group", y = "Mean heart rate (bpm)") +
  theme_classic()
```

```{r q8-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("Start from the ggplot() line provided and add the bar and error-bar layers.")
  }
  geoms   <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  has_bar <- any(c("GeomCol", "GeomBar") %in% geoms)
  has_err <- "GeomErrorbar" %in% geoms
  if (has_bar && has_err) {
    pass("Bars for the means, whiskers for the SE — labeled axes, clean theme. That's a lab-ready figure.")
  }
  if (has_bar && !has_err) {
    fail("Bars are there. Now add geom_errorbar(aes(ymin = mean_hr - se_hr, ymax = mean_hr + se_hr), width = 0.2).")
  }
  fail("Add geom_col() for the bars (then geom_errorbar() for the SE).")
})
```

### Small multiples with `facet_wrap()`

`facet_wrap()` splits one plot into a panel per group. Add `+ facet_wrap(~ group)` to the scatter
below.

```{r q9, exercise=TRUE}
# Split the scatter into one panel per group with facet_wrap(~ group).
ggplot(mammals, aes(x = log_mass, y = heart_rate_bpm)) +
  geom_point() +
  labs(x = "log10 body mass (kg)", y = "Heart rate (bpm)") +
  theme_classic()
```

```{r q9-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("Start from the scatter plot and add + facet_wrap(~ group).")
  }
  if (inherits(.result$facet, "FacetWrap")) {
    pass("Four panels — the mass–heart-rate trend holds within each group, not just across them.")
  }
  fail("Add + facet_wrap(~ group) to split the plot into one panel per group.")
})
```

### A trend line with its equation and R²

A scatter *shows* a relationship; a fitted line *summarizes* it. `geom_smooth(method = "lm")` adds the
least-squares regression line. Two `ggpubr` layers print the numbers right on the figure:
`stat_regline_equation()` writes the line's equation, and `stat_cor(aes(label = ..rr.label..))` writes
its R² — the fraction of the variation the line explains. This is exactly the regression figure you'll
build for the algae experiment, and it's on R Exam 1.

Add three layers to the scatter below: `geom_smooth(method = "lm")`, then `stat_regline_equation()`,
then `stat_cor(aes(label = ..rr.label..))`.

```{r q10, exercise=TRUE}
# Heart rate falls as body size rises. Add the fitted line and its stats:
#   + geom_smooth(method = "lm")
#   + stat_regline_equation()
#   + stat_cor(aes(label = ..rr.label..))
ggplot(mammals, aes(x = log_mass, y = heart_rate_bpm)) +
  geom_point() +
  labs(x = "log10 body mass (kg)", y = "Heart rate (bpm)") +
  theme_classic()
```

```{r q10-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail('Start from the scatter and add + geom_smooth(method = "lm").')
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (any(c("GeomSmooth", "GeomLine") %in% geoms)) {
    pass("There's the trend line with its equation and R² — a complete regression figure.")
  }
  fail('Add + geom_smooth(method = "lm") to draw the fitted line.')
})
```

```{r coach-ui-q10, echo=FALSE}
BisonExplorR::coach_ui("coach_q10", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

### Saving a figure with `ggsave()`

To hand a figure in, you save it to a file. `ggsave()` writes a plot to an image — give it a filename
ending in `.png`, the plot object, and a size. On both R exams you'll save each figure this way and
upload it, so make the pattern automatic now.

The finished regression figure is stored in `p`. Save it as `"mass_hr.png"` with
`ggsave("mass_hr.png", p, width = 16, height = 10, units = "cm")`.

```{r q11, exercise=TRUE}
p <- ggplot(mammals, aes(x = log_mass, y = heart_rate_bpm)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "log10 body mass (kg)", y = "Heart rate (bpm)") +
  theme_classic()

# Add one line: save p as "mass_hr.png" with ggsave(), width = 16,
# height = 10, units = "cm".

```

```{r q11-check}
grade_this({
  code <- paste(.user_code, collapse = "\n")
  if (!grepl("ggsave", code)) {
    fail("Use ggsave() to save the figure to a file.")
  }
  if (!grepl("\\.png", code)) {
    fail('Give ggsave() a filename ending in .png, like "mass_hr.png".')
  }
  pass("Saved — a .png like that is what you'd upload on exam day.")
})
```

---

## You made it

In this module you learned to:

- chain steps with `%>%`
- reshape data with `filter()`, `select()`, `mutate()`, `group_by()`, and `summarise()`
- compute a standard error and turn it into error bars
- build scatter plots, boxplots, faceted plots, and SE bar charts in `ggplot2`
- add a fitted regression line with its equation and R², and save a figure with `ggsave()`

*(Combining two separate tables — joining — comes later, in the cardio unit, once you have data
that actually arrives in more than one file.)*

Next up is **Module 5**, where these figures meet statistics: choosing and running the right test,
then interpreting what it tells you. Everything you wrangled and plotted here is the foundation for
that.
