---
title: "For students"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{For students}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This guide covers everything you need to do BIOL 204 R work: opening tutorials,
getting lab scripts, and practicing for exams. If you are new to R, start at the top
and read straight through — it takes about ten minutes.

You do not need to install R on your own computer. All of this runs in **Posit Cloud**
in a web browser.

## Getting set up

Open the course Posit Cloud workspace using the link on Moodle, then run:

```{r}
library(BisonExplorR)
```

If that produces an error saying the package is not found, tell your instructor rather
than trying to install it yourself — the workspace should already have it.

## Two ways to open a tutorial

Every tutorial exists in two versions, and **they are not interchangeable**.

### The graded version

This is the one that counts. Open it with:

```{r}
launch_graded_tutorial("module5")
```

Run it with no arguments to see the full list:

```{r}
launch_graded_tutorial()
```

The graded copy opens with a **navy banner** reading "Graded copy" and asks for your
student ID. You cannot continue past that first section until you enter it.

### The practice version

Use this to review, or to work through a module a second time without it being recorded:

```{r}
learnr::run_tutorial("05-groups", package = "BisonExplorR")
```

The practice copy opens with an **orange banner** reading "Practice mode". It records
nothing.

### How to tell them apart

| You see | You are in | Your work is |
|---|---|---|
| Navy banner + student ID box | the graded copy | recorded |
| Orange banner, no ID box | a practice copy | not recorded |

**If you open a tutorial for credit and see the orange banner, stop and email your
instructor.** It means something is misconfigured and your work is not being saved.
This is worth thirty seconds of your time — it is much easier to fix before you have
done the whole module.

## Your student ID

Use **the same ID every single time**, in every module, all semester.

The system has no way to know that two different IDs belong to the same person. If you
use your Bucknell email in Module 3 and your ID number in Module 4, those look like two
separate students, and neither one has completed both modules.

If you realize partway through the semester that you have been inconsistent, tell your
instructor — it can be fixed, but only if someone knows.

## How tutorials are graded

Each graded module is worth 10 points:

- **5 points for completion** — did you submit an answer to each exercise?
- **5 points for correctness** — did you eventually get each one right?

**Wrong attempts do not cost you anything.** You can submit an exercise as many times as
you like; what is recorded is whether you ever got it right. Guessing and revising is the
intended way to use these tutorials, not a fallback.

One thing to know: **Run and Submit are different buttons.** Running code shows you the
output, but only Submit records an answer. If you run everything and submit nothing, your
completion score will be zero even though you did all the work.

## The Coach

Some exercises have a **"Stuck? Ask the Coach"** button below them. The Coach is an AI
tutor that reads the code you last submitted and gives you one nudge toward the next step.

It will not give you the answer. That is deliberate — it will name a concept, point at a
function, or ask you a question about your own code, then stop and let you try. If you ask
it for the solution outright, it will decline and offer a hint instead.

Two practical notes:

- **The Coach reads your last *submitted* attempt**, not code you only ran. Submit
  something first, even if you know it is wrong.
- The Coach appears only on the exercises where students most often get stuck, so most
  exercises do not have one. That is normal.

## In-lab scripts

Lab work happens in plain R scripts you edit with a partner. Copy one into your project
with:

```{r}
get_lab("comparing-groups")
```

Run it with no arguments to see what is available:

```{r}
get_lab()
```

The file appears in your Files pane and opens in the editor. It is yours to edit — that
is the point.

If you have already started one and want to begin again from scratch:

```{r}
get_lab("comparing-groups", overwrite = TRUE)
```

That **erases the work you have done in that file**, so only use it if you mean it.

## Practicing for exams

Practice exams use different organisms and different numbers from the real thing, but the
same structure, the same skills, and the same point weighting.

Get the datasets with:

```{r}
get_practice_exam("exam2")
```

This copies the CSV files into your project. That matters more than it looks: on exam day
the data files are already sitting in your project, so your code reads them by plain
filename:

```{r}
turtles <- read.csv("turtle_growth.csv")
```

Practicing the same way means the code you have rehearsed is the code that works under
time pressure.

## Your code library

Keep one script — start it with `get_lab("code-library-starter")` — where you save
working code with a short comment about what it does and when to use it. Add to it every
week rather than at the end.

You are allowed to use it during exams. Students who build one as they go consistently
report it was the single most useful thing they had; students who assemble one the night
before find it is mostly a list of things they cannot remember the purpose of.

## If something goes wrong

**"could not find function"** — you have not run `library(BisonExplorR)` in this session,
or the tutorial's setup did not finish loading. Restart R and try again.

**"object not found"** in an exercise — each exercise runs in its own sandbox, so an
object you created in an earlier exercise is not available. Everything you need is loaded
for you; check the spelling against the dataset table at the top of the section.

**A file is already in your project** — `get_lab()` and `get_practice_exam()` refuse to
overwrite your work by default. Either rename your copy, or pass `overwrite = TRUE` if you
genuinely want to start over.

**The tutorial will not let you continue** — check that you have entered your student ID
in the first section. In the graded copy, the Continue button stays disabled until you do.
