---
title: "Module 10: Joining Data — Combining Tables"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: >
  Real data often arrives in more than one table. Learn the two ways to combine
  tables — stacking (bind_rows / bind_cols) and matching on a key (the joins) —
  and how left, inner, full, and right joins differ.
---

```{css bx-style, echo=FALSE}
/* Sidebar is a progress display, not navigation.                   */
/* allow_skip: false gates the Continue button but NOT the sidebar, */
/* so without this a student can click any topic and land there     */
/* having completed nothing.                                        */
/* Disabling the <li>, not the <a>: the <nav> is a shiny-bound-input */
/* that delegates clicks, so an inert <a> just passes the click up.  */
.topicsList .nav-pills .topic { pointer-events: none; }
.topicsList .nav-pills .topic:not(.current) > .nav-link { opacity: 0.5; }

/* Start Over sits in .topicsFooter, outside .nav-pills, so it stays */
/* clickable. The Previous button is deliberately NOT hidden --      */
/* with the sidebar inert it is the only way back to review.         */
```


```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(BisonExplorR)   # coach_ui() / coach_server()
library(gradethis)
library(dplyr)
library(httr)

gradethis_setup()
tutorial_options(
  exercise.checker = gradethis::grade_learnr,
  exercise.completion = TRUE
)
knitr::opts_chunk$set(echo = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Resolve the student ID no matter which session scope a handler receives.
# VERIFIED 2026-07-25 (learnr 0.11.6): question() blocks are Shiny MODULES, so
# a question_submission handler gets a NAMESPACED session whose input$sid is
# EMPTY. The ID lives on the root session. This fails silently -- rows land in
# the sheet looking well-formed but unattributable, grade_module() drops them,
# and the student loses that score with no warning anywhere. Do not "simplify"
# this back to session$input$sid.
bx_get_sid <- function(session) {
  usable <- function(x) !is.null(x) && length(x) > 0 && nzchar(as.character(x)[1])
  sid <- tryCatch(session$rootScope()$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$userData$bx_sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  NULL   # log_attempt()'s %||% turns this into "unknown"
}

# Is this the GRADED deployment or a practice copy?
# Set BISONEXPLORR_GRADED=true in the deployed app's .Renviron (or set the
# option in its startup). Practice copies launched with run_tutorial() never
# see it, so they default to FALSE.
#
# WHY THE PRACTICE BRANCH SHOWS A VISIBLE BANNER RATHER THAN NOTHING: if this
# flag is ever missing on a deployed graded app, the ID box disappears and
# nothing logs -- every student silently scores zero. A banner makes a
# misconfigured deploy announce itself on the first screen instead of three
# weeks later in the response sheet.
bx_is_graded <- function() {
  isTRUE(getOption(
    "BisonExplorR.graded",
    identical(tolower(Sys.getenv("BISONEXPLORR_GRADED")), "true")
  ))
}

log_attempt <- function(student_id, exercise_id, correct) {
  # Practice copies do not log. Without this, practice attempts POST with
  # student_id = "unknown" and fill the response sheet with orphan rows
  # that look like a bug in grade_module().
  if (!bx_is_graded()) return(invisible(NULL))

  tryCatch({
    httr::POST(
      url = "https://docs.google.com/forms/d/e/1FAIpQLSfmm3cUcoaWWYBb0zk7rbowSpgUL13nQc-Dxn7K8zQ30GdykA/formResponse",
      body = list(
        "entry.1413675438" = as.character(Sys.time()),
        "entry.1257984687" = as.character(student_id %||% "unknown"),
        "entry.1315894547" = "module10",
        "entry.1172625272" = as.character(exercise_id %||% "unknown"),
        "entry.1835356043" = as.character(correct %||% "NA")
      ),
      encode = "form"
    )
  }, error = function(e) message("Logging failed: ", e$message))
}

# ---- Shared teaching tables (small on purpose so you can see every row) -----
# One row per field site:
sites <- data.frame(
  site_id = c("P1", "P2", "P3", "P4"),
  habitat = c("forest", "forest", "meadow", "wetland"),
  stringsAsFactors = FALSE
)

# Frog counts, but NOT every site was surveyed, and one survey (P5) is from a
# site that isn't in `sites` at all:
surveys <- data.frame(
  site_id   = c("P1", "P2", "P3", "P5"),
  species   = c("wood frog", "wood frog", "wood frog", "wood frog"),
  abundance = c(12, 8, 3, 5),
  stringsAsFactors = FALSE
)

# Two survey rounds with the SAME columns (for stacking):
spring <- data.frame(site_id = c("P1", "P2", "P3"), abundance = c(12, 8, 3),
                     stringsAsFactors = FALSE)
fall   <- data.frame(site_id = c("P1", "P2", "P3"), abundance = c(15, 6, 4),
                     stringsAsFactors = FALSE)

# GPS coordinates for the four sites, in the SAME ORDER as `sites` (for gluing
# columns side by side). Notice there is no site_id column here:
coords <- data.frame(lat = c(40.95, 40.96, 40.94, 40.97),
                     lon = c(-76.88, -76.89, -76.87, -76.90),
                     stringsAsFactors = FALSE)
```

```{r event-setup, context="server-start"}
event_register_handler("exercise_result", function(session, event, data) {
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = data$feedback$correct
  )
})
```

```{r coach-server, context="server"}
# One coach_server() per coached exercise. The third argument must match a
# key in the server-side exercise_registry.R ("moduleN-qM", HYPHENATED) --
# an unknown key does not error, it silently returns generic context.
# The Shiny id must match the coach_ui() call below it, or the card renders
# and never responds.
BisonExplorR::coach_server("coach_q1", shiny::reactive(learnr::get_tutorial_state("q1")), "module10-q1")
BisonExplorR::coach_server("coach_q4", shiny::reactive(learnr::get_tutorial_state("q4")), "module10-q4")
```

## Before you begin

```{r student-id, echo=FALSE}
uiOutput("sid_ui")
```

```{r sid-server, context="server"}
# The ID box exists ONLY in the graded deployment. See bx_is_graded() in setup.
output$sid_ui <- renderUI({
  if (!bx_is_graded()) {
    return(shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #C05621; background:#FDF6EC;"),
      shiny::strong("Practice mode."),
      " Nothing here is recorded and this copy does not count for credit.",
      " To do this module for credit, run ",
      shiny::tags$code('launch_graded_tutorial("module10")'),
      " instead."
    ))
  }
  shiny::tagList(
    shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #1F3864; background:#F2F4F7;"),
      shiny::strong("Graded copy."),
      " Your work on this module is recorded, so enter your student ID before",
      " you begin. Use the same ID every time \u2014 a different ID reads as a",
      " different student and your work will not be combined."
    ),
    shiny::textInput("sid",
      "Student ID (the same one you used in Module 1):", value = ""),
    shiny::div(id = "sid_warn",
      style = "color:#C05621; font-weight:600; margin:-6px 0 8px 0;",
      "Enter your ID to continue."),
    # Client-side gate on the section's Continue button. This prevents the
    # ACCIDENTAL case (a student who forgets), not a determined one -- anyone
    # with devtools can re-enable the button. The real protection against
    # unattributed rows is that grade_module() drops blank IDs.
    shiny::tags$script(shiny::HTML("
      (function () {
        function apply() {
          var sid = document.getElementById('sid');
          if (!sid) return;
          var sec = sid.closest('.section');
          if (!sec) return;
          var btns = sec.querySelectorAll('.topicActions button, .continueButton');
          var ok = sid.value.trim().length > 0;
          btns.forEach(function (b) {
            b.disabled = !ok;
            b.style.opacity = ok ? '' : '0.45';
            b.style.cursor  = ok ? '' : 'not-allowed';
          });
          var w = document.getElementById('sid_warn');
          if (w) w.style.display = ok ? 'none' : 'block';
        }
        document.addEventListener('input', function (e) {
          if (e.target && e.target.id === 'sid') apply();
        });
        // learnr renders sections progressively, so the Continue button may not
        // exist yet when this script first runs.
        new MutationObserver(apply)
          .observe(document.body, { childList: true, subtree: true });
        apply();
      })();
    "))
  )
})

# Cache the ID on userData as a second, independent path to it. bx_get_sid()
# checks rootScope() first and falls back to this.
shiny::observeEvent(input$sid, {
  if (!is.null(input$sid) && nzchar(input$sid)) {
    session$userData$bx_sid <- input$sid
  }
}, ignoreInit = TRUE)
```


## Why tables come apart

In the field you rarely record everything in one sheet. You keep a **site list**
(where the sites are, what habitat they're in) separate from your **survey
data** (what you counted, and when). That's good practice — but eventually you
need them back together.

There are two fundamentally different ways to combine tables, and telling them
apart is the whole game:

| You want to… | Tables share… | Tool |
|---|---|---|
| **stack** tables on top of each other (more rows) | the same columns | `bind_rows()` |
| **glue** tables side by side (more columns) | the same rows, in order | `bind_cols()` |
| **match** rows by a shared key | a key column (e.g. `site_id`) | a **join** |

Meet our two tables. Run these and look at them:

```{r peek-sites, exercise=TRUE}
sites
```

```{r peek-surveys, exercise=TRUE}
surveys
```

Notice the mismatch — it will matter a lot in a minute. **`sites`** lists P1–P4.
**`surveys`** has P1, P2, P3, and **P5**. So **P4 was never surveyed**, and **P5
was surveyed but isn't in your site list**.

## Stacking rows: `bind_rows()` {data-allow-skip="true"}

When two tables describe the *same kind of thing* with the *same columns* — like
two rounds of the same survey — you stack them into one longer table with
`bind_rows()`. It lines columns up **by name**, so the column order doesn't have
to match.

Worked example — one spring round and one fall round:

```{r bindrows-demo, exercise=TRUE}
spring
fall
```

`bind_rows()` can add an `.id` column to record which table each row came from —
handy for keeping the seasons straight:

```{r bindrows-demo2, exercise=TRUE}
bind_rows(spring = spring, fall = fall, .id = "season")
```

### Your turn

```{r q1, exercise=TRUE}
# Stack `spring` and `fall` into one table (no .id needed here).
bind_rows()
```

```{r q1-check}
grade_this({
  correct <- bind_rows(spring, fall)
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct),
                       check.attributes = FALSE))) {
    pass("Six rows — the three spring plus the three fall. Stacking keeps every row.")
  }
  fail("Not quite. Pass both tables to bind_rows(): bind_rows(spring, fall).")
})
```

```{r coach-ui-q1, echo=FALSE}
BisonExplorR::coach_ui("coach_q1", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

## Gluing columns: `bind_cols()`

`bind_cols()` sets tables **side by side**. It does **not** look at any key — it
matches rows purely by **position**: row 1 with row 1, row 2 with row 2, and so
on. That makes it powerful but dangerous.

Here `coords` has GPS points listed in the same order as `sites`:

```{r bindcols-demo, exercise=TRUE}
bind_cols(sites, coords)
```

```{r bindcols-safety, echo=FALSE}
question(
  "What happens if `coords` were sorted in a DIFFERENT order than `sites`?",
  answer("Every site would get the wrong coordinates, with no warning.",
         correct = TRUE,
         message = "`bind_cols()` trusts row order blindly. When a shared key exists, a join is safer."),
  answer("R would automatically match them by site_id.",
         message = "No — bind_cols() ignores keys entirely. It only sees row position."),
  answer("R would throw an error.",
         message = "Only if the row COUNTS differ. Same count but wrong order fails silently."),
  allow_retry = TRUE
)
```

The lesson: reach for `bind_cols()` only when you're certain the rows line up.
Whenever the tables share a key column, use a **join** instead — that's next.

## Matching on a key: the joins {data-allow-skip="true"}

A **join** combines two tables by matching a shared **key** — here, `site_id`.
The four joins differ in **one thing only**: *which rows do you keep when a key
appears in one table but not the other?*

Keep this map in view:

| Join | Keeps… | Rows here |
|---|---|---|
| `inner_join(x, y)` | only keys in **both** tables | P1, P2, P3 |
| `left_join(x, y)`  | **all of `x`**, matched where possible | P1, P2, P3, P4 |
| `right_join(x, y)` | **all of `y`**, matched where possible | P1, P2, P3, P5 |
| `full_join(x, y)`  | **everything** from both | P1, P2, P3, P4, P5 |

**Course convention:** always name the key explicitly with `by = "..."`. It makes
your intent clear and guards against joining on the wrong column by accident.

### `left_join()` — keep all of the left table

This is the workhorse. "Start from `sites`, and attach survey data where it
exists." Sites with no survey (P4) are **kept**, with `NA` in the survey columns.

```{r q2, exercise=TRUE}
# Attach the survey data to every site. Fill in the key column name.
sites %>%
  left_join(surveys, by = "____")
```

```{r q2-check}
grade_this({
  correct <- left_join(sites, surveys, by = "site_id")
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct),
                       check.attributes = FALSE))) {
    pass("Four rows — all of `sites`. Look at P4: it stayed, but its survey columns are NA because it was never surveyed.")
  }
  fail('Not quite. The shared key is "site_id": left_join(surveys, by = "site_id").')
})
```

### `inner_join()` — keep only matches

"Only sites that were actually surveyed." P4 (no survey) and P5 (no site record)
both **drop out**.

```{r q3, exercise=TRUE}
# Keep only sites that appear in BOTH tables.
sites %>%
  inner_join(surveys, by = "____")
```

```{r q3-check}
grade_this({
  correct <- inner_join(sites, surveys, by = "site_id")
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct),
                       check.attributes = FALSE))) {
    pass("Three rows — P1, P2, P3. An inner join keeps only keys present in both tables.")
  }
  fail('Not quite. Use inner_join(surveys, by = "site_id"). You should get three rows.')
})
```

### `full_join()` — keep everything

"Show me every site AND every survey, matched where possible." Nothing is lost —
P4 (no survey) and P5 (no site) both appear, each with `NA` where the other table
had nothing.

```{r q4, exercise=TRUE}
# Keep every row from both tables.
sites %>%
  full_join(surveys, by = "____")
```

```{r q4-check}
grade_this({
  correct <- full_join(sites, surveys, by = "site_id")
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct),
                       check.attributes = FALSE))) {
    pass("Five rows — the union of both tables. P4 has NA survey data; P5 has NA habitat.")
  }
  fail('Not quite. Use full_join(surveys, by = "site_id"). You should get five rows.')
})
```

```{r coach-ui-q4, echo=FALSE}
BisonExplorR::coach_ui("coach_q4", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

### `right_join()` — the mirror image

`right_join(x, y)` keeps all of **`y`**. It's just `left_join()` with the tables
swapped, so most people pick one and stick with `left_join()`. Run this to see
that P5 (survey with no site) is the one that's kept, and P4 drops:

```{r rightjoin-demo, exercise=TRUE}
right_join(sites, surveys, by = "site_id")
```

## Keeping or dropping rows: filtering joins

The joins above **add columns** from the other table. Two special joins add
**no columns at all** — they only use the second table to decide which rows of
the first to keep. They're perfect for asking "which of my sites *do* / *don't*
have survey data?"

- `semi_join(x, y)` — keep rows of `x` that **have** a match in `y`.
- `anti_join(x, y)` — keep rows of `x` that have **no** match in `y`.

```{r semijoin-demo, exercise=TRUE}
# Sites that WERE surveyed (columns of `sites` only — nothing added):
semi_join(sites, surveys, by = "site_id")
```

### Your turn

```{r q5, exercise=TRUE}
# Find the site(s) that were NEVER surveyed — rows of `sites` with no match.
anti_join(sites, surveys, by = "____")
```

```{r q5-check}
grade_this({
  correct <- anti_join(sites, surveys, by = "site_id")
  if (isTRUE(all.equal(as.data.frame(.result), as.data.frame(correct),
                       check.attributes = FALSE))) {
    pass("Just P4 — the wetland site that was never surveyed. anti_join() is the fastest way to find gaps in your data.")
  }
  fail('Not quite. Use anti_join(surveys, by = "site_id"). One row (P4) should remain.')
})
```

## Which combine do I reach for?

```{r decide, echo=FALSE}
question(
  "You have a `sites` table and a `surveys` table sharing `site_id`. You want ONE table with every site listed, plus its survey counts where they exist, and blanks where a site wasn't surveyed. Which tool?",
  answer("left_join(sites, surveys, by = \"site_id\")", correct = TRUE,
         message = "You want all of the left table (sites) with survey data attached where available."),
  answer("bind_rows(sites, surveys)",
         message = "bind_rows() stacks rows and needs matching columns; these tables have different columns and a shared key, so it's a join."),
  answer("inner_join(sites, surveys, by = \"site_id\")",
         message = "Inner join would DROP unsurveyed sites — but you said you want every site listed."),
  answer("bind_cols(sites, surveys)",
         message = "bind_cols() matches by position and ignores site_id; the tables don't even have the same rows."),
  allow_retry = TRUE
)
```

```{r decide2, echo=FALSE}
question(
  "You ran the same survey protocol in two years and each year's data is in its own table with identical columns. You want one long table for analysis. Which tool?",
  answer("bind_rows()", correct = TRUE,
         message = "Same columns, you just want more rows. Add an .id column to remember which year."),
  answer("full_join()",
         message = "A join matches on a key; here you simply want to stack identical-shaped tables."),
  answer("bind_cols()",
         message = "That would set the years side by side as extra columns, not stack them into more rows."),
  allow_retry = TRUE
)
```

## Where you landed

You can now combine tables two ways and choose between them:

- **Stack** with `bind_rows()` (more rows, same columns) — line up by name.
- **Glue** with `bind_cols()` (more columns, same rows) — matched by position, so
  only when you're sure the order is identical.
- **Join** on a key when tables share an ID column:
  - `left_join()` keeps all of the left table (the everyday default),
  - `inner_join()` keeps only matches,
  - `full_join()` keeps everything,
  - `right_join()` keeps all of the right table.
- **Filtering joins** `semi_join()` / `anti_join()` keep or drop left rows by
  whether they match — great for finding the gaps in your data.

Always write the key explicitly: `by = "site_id"`. In the cardio unit you'll use
`left_join()` to attach subject information to heart-rate measurements — the same
move you practiced here.
