---
title: "Module 5: Comparing Groups — t-tests and ANOVA"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: >
  Ask whether two or more groups differ. Run a t-test and a one-way ANOVA,
  use a post-hoc test to see which groups differ, and show the comparison.
---

```{css bx-style, echo=FALSE}
/* Sidebar is a progress display, not navigation.                   */
/* allow_skip: false gates the Continue button but NOT the sidebar, */
/* so without this a student can click any topic and land there     */
/* having completed nothing.                                        */
/* Disabling the <li>, not the <a>: the <nav> is a shiny-bound-input */
/* that delegates clicks, so an inert <a> just passes the click up.  */
.topicsList .nav-pills .topic { pointer-events: none; }
.topicsList .nav-pills .topic:not(.current) > .nav-link { opacity: 0.5; }

/* Start Over sits in .topicsFooter, outside .nav-pills, so it stays */
/* clickable. The Previous button is deliberately NOT hidden --      */
/* with the sidebar inert it is the only way back to review.         */
```


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(httr)
library(ggplot2)
library(shiny)
library(BisonExplorR)   # coach_ui() / coach_server()
gradethis::gradethis_setup()
tutorial_options(
  exercise.checker = gradethis::grade_learnr,
  exercise.completion = TRUE
)
knitr::opts_chunk$set(echo = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Resolve the student ID no matter which session scope a handler receives.
# VERIFIED 2026-07-25 (learnr 0.11.6): question() blocks are Shiny MODULES, so
# a question_submission handler gets a NAMESPACED session whose input$sid is
# EMPTY. The ID lives on the root session. This fails silently -- rows land in
# the sheet looking well-formed but unattributable, grade_module() drops them,
# and the student loses that score with no warning anywhere. Do not "simplify"
# this back to session$input$sid.
bx_get_sid <- function(session) {
  usable <- function(x) !is.null(x) && length(x) > 0 && nzchar(as.character(x)[1])
  sid <- tryCatch(session$rootScope()$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$userData$bx_sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  NULL   # log_attempt()'s %||% turns this into "unknown"
}

# Is this the GRADED deployment or a practice copy?
# Set BISONEXPLORR_GRADED=true in the deployed app's .Renviron (or set the
# option in its startup). Practice copies launched with run_tutorial() never
# see it, so they default to FALSE.
#
# WHY THE PRACTICE BRANCH SHOWS A VISIBLE BANNER RATHER THAN NOTHING: if this
# flag is ever missing on a deployed graded app, the ID box disappears and
# nothing logs -- every student silently scores zero. A banner makes a
# misconfigured deploy announce itself on the first screen instead of three
# weeks later in the response sheet.
bx_is_graded <- function() {
  isTRUE(getOption(
    "BisonExplorR.graded",
    identical(tolower(Sys.getenv("BISONEXPLORR_GRADED")), "true")
  ))
}

log_attempt <- function(student_id, exercise_id, correct) {
  # Practice copies do not log. Without this, practice attempts POST with
  # student_id = "unknown" and fill the response sheet with orphan rows
  # that look like a bug in grade_module().
  if (!bx_is_graded()) return(invisible(NULL))

  tryCatch({
    httr::POST(
      url = "https://docs.google.com/forms/d/e/1FAIpQLSfmm3cUcoaWWYBb0zk7rbowSpgUL13nQc-Dxn7K8zQ30GdykA/formResponse",
      body = list(
        "entry.1413675438" = as.character(Sys.time()),
        "entry.1257984687" = as.character(student_id %||% "unknown"),
        "entry.1315894547" = "module5",
        "entry.1172625272" = as.character(exercise_id %||% "unknown"),
        "entry.1835356043" = as.character(correct %||% "NA")
      ),
      encode = "form"
    )
  }, error = function(e) message("Logging failed: ", e$message))
}

# Shared data — must live in the setup chunk (learnr sandboxes each exercise).
# Read from the packaged CSV rather than generated with set.seed(): the numbers
# quoted in the pass messages below were verified against THIS file, so a change
# in R's RNG cannot silently invalidate them.
algae <- read.csv(
  system.file("extdata", "algae_beads.csv", package = "BisonExplorR"),
  stringsAsFactors = FALSE
)

# House style: factors are pre-set here, never with inline as.factor().
algae$species        <- factor(algae$species)
algae$nutrient_level <- factor(algae$nutrient_level,
                               levels = c("low", "medium", "high"))
```

```{r event-setup, context="server-start"}
# Code exercises. Correctness lives on data$feedback$correct here.
event_register_handler("exercise_result", function(session, event, data) {
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = data$feedback$correct
  )
})

# Multiple-choice items. This is what makes mc1-mc4 gradeable: without it
# question() blocks emit question_submission, nothing logs them, and
# grade_module() has no rows to score against.
#
# NOTE: the payload shape differs from exercise_result -- correctness is on
# data$correct, NOT data$feedback$correct. That field name has not been
# verified against the installed learnr, and a wrong guess fails SILENTLY
# (every MC logs NA forever), so both are tried and the probe confirms it.
event_register_handler("question_submission", function(session, event, data) {
  correct_flag <- data$correct
  if (is.null(correct_flag) || length(correct_flag) == 0) {
    correct_flag <- data$feedback$correct
  }
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = correct_flag
  )
})
```

```{r coach-server, context="server"}
# Coached: q1 (first stats code they write) and q3 (composing two functions).
# q2 mirrors q1's pattern and q4 is a single-blank fill — the inline hint is
# enough there, and a coach would only add noise and API cost.
BisonExplorR::coach_server("coach_q1", shiny::reactive(learnr::get_tutorial_state("q1")), "module5-q1")
BisonExplorR::coach_server("coach_q3", shiny::reactive(learnr::get_tutorial_state("q3")), "module5-q3")
```


## Before you start

```{r student-id, echo=FALSE}
uiOutput("sid_ui")
```

```{r sid-server, context="server"}
# The ID box exists ONLY in the graded deployment. See bx_is_graded() in setup.
output$sid_ui <- renderUI({
  if (!bx_is_graded()) {
    return(shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #C05621; background:#FDF6EC;"),
      shiny::strong("Practice mode."),
      " Nothing here is recorded and this copy does not count for credit.",
      " To do this module for credit, run ",
      shiny::tags$code('launch_graded_tutorial("module5")'),
      " instead."
    ))
  }
  shiny::tagList(
    shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #1F3864; background:#F2F4F7;"),
      shiny::strong("Graded copy."),
      " Your work on this module is recorded, so enter your student ID before",
      " you begin. Use the same ID every time \u2014 a different ID reads as a",
      " different student and your work will not be combined."
    ),
    shiny::textInput("sid",
      "Student ID (the same one you used in Module 1):", value = ""),
    shiny::div(id = "sid_warn",
      style = "color:#C05621; font-weight:600; margin:-6px 0 8px 0;",
      "Enter your ID to continue."),
    # Client-side gate on the section's Continue button. This prevents the
    # ACCIDENTAL case (a student who forgets), not a determined one -- anyone
    # with devtools can re-enable the button. The real protection against
    # unattributed rows is that grade_module() drops blank IDs.
    shiny::tags$script(shiny::HTML("
      (function () {
        function apply() {
          var sid = document.getElementById('sid');
          if (!sid) return;
          var sec = sid.closest('.section');
          if (!sec) return;
          var btns = sec.querySelectorAll('.topicActions button, .continueButton');
          var ok = sid.value.trim().length > 0;
          btns.forEach(function (b) {
            b.disabled = !ok;
            b.style.opacity = ok ? '' : '0.45';
            b.style.cursor  = ok ? '' : 'not-allowed';
          });
          var w = document.getElementById('sid_warn');
          if (w) w.style.display = ok ? 'none' : 'block';
        }
        document.addEventListener('input', function (e) {
          if (e.target && e.target.id === 'sid') apply();
        });
        // learnr renders sections progressively, so the Continue button may not
        // exist yet when this script first runs.
        new MutationObserver(apply)
          .observe(document.body, { childList: true, subtree: true });
        apply();
      })();
    "))
  )
})

# Cache the ID on userData as a second, independent path to it. bx_get_sid()
# checks rootScope() first and falls back to this.
shiny::observeEvent(input$sid, {
  if (!is.null(input$sid) && nzchar(input$sid)) {
    session$userData$bx_sid <- input$sid
  }
}, ignoreInit = TRUE)
```


## The question this module answers

You have a response you measured and groups you want to compare. *Do the groups
differ by more than chance?* That is what a t-test and an ANOVA are for — the
only difference is **how many groups**.

Everything here uses **`algae`**: 60 vials of alginate-encapsulated algae beads.
Each vial holds one of two species, was grown at one of three nutrient levels,
and had its **CO2 drawdown** measured — how much CO2 the beads pulled out of the
water. More drawdown means more photosynthesis.

```{r peek, exercise=TRUE}
head(algae)
```

This is the same response variable you will meet again in Module 6
(`co2_drawdown_ppm`). That is deliberate: one measurement, asked three different
questions. What changes between modules is not the biology but the **kind of
predictor** you put on the right of the tilde.

### The map: which test, when

Keep this table in view — it is the same in every statistics module. The last
column tells you where each test is taught.

| Biological question | Predictor(s) | Test | R function | Module |
|---|---|---|---|---|
| Do **two** groups differ? | one factor, 2 levels | t-test | `t.test()` | 5 |
| Do **three or more** groups differ? | one factor, 3+ levels | one-way ANOVA | `aov()` + `summary()` | 5 |
| Does a **continuous** variable predict the response? | one continuous | linear regression | `lm()` + `summary()` | 6 |
| Do groups differ across **two factors**? | two factors | two-way ANOVA | `aov()` + `summary()` | 7 |
| Do groups differ **after controlling for** a continuous variable? | factor + covariate | ANCOVA | `aov()` + `summary()` | 8 |

**Where you are:** Module 5 — one grouping factor, two or more groups.

### Pick the test

```{r mc1, echo=FALSE}
question(
  "You want to know whether CO2 drawdown differs between the two algal species. Which test?",
  answer("t-test", correct = TRUE, message = "Exactly two groups — a t-test."),
  answer("One-way ANOVA", message = "ANOVA is for three or more groups. Species has two."),
  answer("TukeyHSD", message = "Tukey is a follow-up to ANOVA, not a stand-alone two-group test."),
  allow_retry = TRUE
)
```

## Two groups: the t-test {data-allow-skip="true"}

The t-test formula reads `response ~ group`.

### Exercise q1

Run a t-test comparing `co2_drawdown_ppm` between the two levels of `species`.

```{r q1, exercise=TRUE}
t.test(___ ~ ___, data = algae)
```

```{r q1-hint}
# Response on the left of ~, grouping factor on the right.
t.test(co2_drawdown_ppm ~ species, data = algae)
```

```{r q1-solution}
t.test(co2_drawdown_ppm ~ species, data = algae)
```

```{r q1-check}
grade_this({
  if (inherits(.result, "htest") && grepl("species", .result$data.name)) {
    pass("Look at the p-value, then the group means at the bottom of the output.")
  }
  fail("Not quite. Use t.test(co2_drawdown_ppm ~ species, data = algae).")
})
```

```{r coach-ui-q1, echo=FALSE}
BisonExplorR::coach_ui("coach_q1", "Stuck? Ask the Coach")
```

### Read it

The output gave `t = -2.36`, `p = 0.022`, with mean drawdown of **34.3 ppm** for
*Chlorella* and **39.0 ppm** for *Scenedesmus*.

```{r mc2, echo=FALSE}
question(
  "At the 0.05 level, what does this say?",
  answer("Scenedesmus drew down significantly more CO2 than Chlorella in this experiment.", correct = TRUE),
  answer("There is no significant difference.", message = "p = 0.022 is below 0.05."),
  answer("Scenedesmus is the better species for every purpose.", message = "The test compares one response under one set of conditions — it does not rank the species generally."),
  allow_retry = TRUE
)
```

## Three or more groups: one-way ANOVA

Nutrient level has three groups, so a t-test no longer fits — you run a
**one-way ANOVA**. The course convention is `aov()` wrapped in `summary()`:
`aov()` fits the model, `summary()` prints the F value and p-value.

### Exercise q2

Fit a one-way ANOVA of `co2_drawdown_ppm` by `nutrient_level` and print the summary.

```{r q2, exercise=TRUE}
summary(aov(___ ~ ___, data = algae))
```

```{r q2-hint}
summary(aov(co2_drawdown_ppm ~ nutrient_level, data = algae))
```

```{r q2-solution}
summary(aov(co2_drawdown_ppm ~ nutrient_level, data = algae))
```

```{r q2-check}
grade_this({
  if (inherits(.result, "summary.aov") && any(grepl("nutrient", rownames(.result[[1]])))) {
    pass("F = 23.62 on 2 and 57 degrees of freedom, p = 3.4e-08. Something differs among the three nutrient levels — but ANOVA won't tell you which. That's the next step.")
  }
  fail("Not quite. Wrap aov(co2_drawdown_ppm ~ nutrient_level, data = algae) in summary().")
})
```

```{r mc3, echo=FALSE}
question(
  "Why run one ANOVA instead of three separate t-tests (low vs. medium, low vs. high, medium vs. high)?",
  answer("Each extra test adds another chance of a false positive, so three tests inflate the overall error rate. ANOVA asks the question once.", correct = TRUE),
  answer("Three t-tests would give exactly the same answer, so ANOVA just saves typing.", message = "It saves typing, but that isn't the reason. Think about what happens to your chance of a false positive each time you run another test."),
  answer("t-tests only work on two groups, so they cannot be run here at all.", message = "You could physically run them on any pair. The problem is what repeating the test does to your error rate."),
  answer("ANOVA is more sensitive, so it always finds a difference when one exists.", message = "No test is guaranteed to detect a real difference; that depends on effect size and sample size."),
  allow_retry = TRUE
)
```

## Which groups differ? TukeyHSD {data-allow-skip="true"}

A significant ANOVA means *at least one* group stands out, not which one. Run
**Tukey's Honest Significant Difference** on the fitted model to compare every
pair.

### Exercise q3

Write this call yourself: run `TukeyHSD()` on an `aov()` model of
`co2_drawdown_ppm` by `nutrient_level`. Note that the post-hoc test runs on the
**fitted model**, not on the summary of it.

```{r q3, exercise=TRUE}

```

```{r q3-hint}
# TukeyHSD() takes the fitted aov() object directly:
# TukeyHSD(aov(response ~ factor, data = algae))
```

```{r q3-solution}
TukeyHSD(aov(co2_drawdown_ppm ~ nutrient_level, data = algae))
```

```{r q3-check}
grade_this({
  if (inherits(.result, "TukeyHSD") && "nutrient_level" %in% names(.result)) {
    pass("Read the `p adj` column: the high group differs from both others (p < 0.0001), but medium vs. low is not significant (p = 0.76, a difference of only 1.4 ppm). The omnibus ANOVA could never have told you that.")
  }
  fail("Not quite. Pass aov(co2_drawdown_ppm ~ nutrient_level, data = algae) to TukeyHSD().")
})
```

```{r coach-ui-q3, echo=FALSE}
BisonExplorR::coach_ui("coach_q3", "Stuck? Ask the Coach")
```

```{r mc4, echo=FALSE}
question(
  "The ANOVA was significant, but Tukey shows medium vs. low is not (p = 0.76). What does that mean?",
  answer("The overall nutrient effect is driven mainly by the high group; medium and low are statistically indistinguishable.", correct = TRUE),
  answer("The ANOVA was wrong.", message = "Both are correct — they answer different questions. ANOVA asks 'any difference?'; Tukey asks 'which pairs?'"),
  answer("All three nutrient levels differ from each other.", message = "Tukey shows medium and low do not differ significantly."),
  allow_retry = TRUE
)
```

## Show the comparison

Numbers convince you; a figure convinces a reader. A boxplot is the standard way
to show a group comparison.

### Exercise q4

Build a boxplot of `co2_drawdown_ppm` by `nutrient_level`. Label the axes and
finish with `theme_classic()`.

```{r q4, exercise=TRUE}
ggplot(algae, aes(x = nutrient_level, y = co2_drawdown_ppm)) +
  geom_boxplot() +
  labs(x = "Nutrient level", y = "CO2 drawdown (ppm)") +
  ___
```

```{r q4-hint}
# theme_classic()
```

```{r q4-solution}
ggplot(algae, aes(x = nutrient_level, y = co2_drawdown_ppm)) +
  geom_boxplot() +
  labs(x = "Nutrient level", y = "CO2 drawdown (ppm)") +
  theme_classic()
```

```{r q4-check}
grade_this({
  if (inherits(.result, "ggplot")) {
    geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
    if (any(geoms == "GeomBoxplot")) {
      pass("The high box sits clearly above the other two, which sit almost level with each other — the same story Tukey told, now legible at a glance.")
    }
  }
  fail("Not quite. Use geom_boxplot() and finish with theme_classic().")
})
```

## Where you landed

- **Two groups ->** `t.test(response ~ group)`
- **Three or more ->** `summary(aov(response ~ group))`, then `TukeyHSD()` to find which pairs differ
- **Show it** with a boxplot

Next module keeps a single predictor but changes its **type**: instead of a
grouping factor, a continuous measurement. That is **linear regression**.
