---
title: "Module 7: Two-way ANOVA"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: >
  Add a second grouping factor. Fit a two-way ANOVA, read the interaction first,
  and show both factors at once with a grouped boxplot and an interaction plot.
---

```{css bx-style, echo=FALSE}
/* Sidebar is a progress display, not navigation.                   */
/* allow_skip: false gates the Continue button but NOT the sidebar, */
/* so without this a student can click any topic and land there     */
/* having completed nothing.                                        */
/* Disabling the <li>, not the <a>: the <nav> is a shiny-bound-input */
/* that delegates clicks, so an inert <a> just passes the click up.  */
.topicsList .nav-pills .topic { pointer-events: none; }
/* learnr paints a topicProgress.png sprite as the <li> background. */
/* With the sidebar inert it reads as clutter.                     */
.topicsList .nav-pills .topic { background-image: none; }
/* 0.7 not 0.5: at 0.5 the dimmed text is 2.80:1, below the 4.5:1  */
/* WCAG AA floor. 0.7 measures 4.76:1 and still reads as dimmed.   */
.topicsList .nav-pills .topic:not(.current) > .nav-link { opacity: 0.7; }

/* Start Over sits in .topicsFooter, outside .nav-pills, so it stays */
/* clickable. The Previous button is deliberately NOT hidden --      */
/* with the sidebar inert it is the only way back to review.         */
```


```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(BisonExplorR)   # coach_ui() / coach_server()
library(gradethis)
library(httr)
library(ggplot2)
gradethis::gradethis_setup()
tutorial_options(
  exercise.checker = gradethis::grade_learnr,
  exercise.completion = TRUE
)
knitr::opts_chunk$set(echo = FALSE)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Resolve the student ID no matter which session scope a handler receives.
# VERIFIED 2026-07-25 (learnr 0.11.6): question() blocks are Shiny MODULES, so
# a question_submission handler gets a NAMESPACED session whose input$sid is
# EMPTY. The ID lives on the root session. This fails silently -- rows land in
# the sheet looking well-formed but unattributable, grade_module() drops them,
# and the student loses that score with no warning anywhere. Do not "simplify"
# this back to session$input$sid.
bx_get_sid <- function(session) {
  usable <- function(x) !is.null(x) && length(x) > 0 && nzchar(as.character(x)[1])
  sid <- tryCatch(session$rootScope()$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$userData$bx_sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  sid <- tryCatch(session$input$sid, error = function(e) NULL)
  if (usable(sid)) return(sid)
  NULL   # log_attempt()'s %||% turns this into "unknown"
}

# Is this the GRADED deployment or a practice copy?
# Set BISONEXPLORR_GRADED=true in the deployed app's .Renviron (or set the
# option in its startup). Practice copies launched with run_tutorial() never
# see it, so they default to FALSE.
#
# WHY THE PRACTICE BRANCH SHOWS A VISIBLE BANNER RATHER THAN NOTHING: if this
# flag is ever missing on a deployed graded app, the ID box disappears and
# nothing logs -- every student silently scores zero. A banner makes a
# misconfigured deploy announce itself on the first screen instead of three
# weeks later in the response sheet.
bx_is_graded <- function() {
  isTRUE(getOption(
    "BisonExplorR.graded",
    identical(tolower(Sys.getenv("BISONEXPLORR_GRADED")), "true")
  ))
}

log_attempt <- function(student_id, exercise_id, correct) {
  # Practice copies do not log. Without this, practice attempts POST with
  # student_id = "unknown" and fill the response sheet with orphan rows
  # that look like a bug in grade_module().
  if (!bx_is_graded()) return(invisible(NULL))

  tryCatch({
    httr::POST(
      url = "https://docs.google.com/forms/d/e/1FAIpQLSfmm3cUcoaWWYBb0zk7rbowSpgUL13nQc-Dxn7K8zQ30GdykA/formResponse",
      body = list(
        "entry.1413675438" = as.character(Sys.time()),
        "entry.1257984687" = as.character(student_id %||% "unknown"),
        "entry.1315894547" = "module7",
        "entry.1172625272" = as.character(exercise_id %||% "unknown"),
        "entry.1835356043" = as.character(correct %||% "NA")
      ),
      encode = "form"
    )
  }, error = function(e) message("Logging failed: ", e$message))
}

# Shared data — must live in the setup chunk (learnr sandboxes each exercise).
# Read from the packaged CSV rather than generated with set.seed(): the numbers
# quoted in the pass messages below were verified against THIS file, so a change
# in R's RNG cannot silently invalidate them.
#
# Same dataset as Module 5, deliberately. There students ran a t-test on species
# and a one-way ANOVA on nutrient_level SEPARATELY; here the two factors are
# fitted together, which is the whole point of the module.
algae <- read.csv(
  system.file("extdata", "algae_beads.csv", package = "BisonExplorR"),
  stringsAsFactors = FALSE
)

# House style: factors are pre-set here, never with inline as.factor().
# nutrient_level needs an explicit level order — alphabetical would print
# high, low, medium, which reads as nonsense on an axis.
algae$species        <- factor(algae$species)
algae$nutrient_level <- factor(algae$nutrient_level,
                               levels = c("low", "medium", "high"))

# Group means for the interaction plot (q3).
algae_means <- aggregate(co2_drawdown_ppm ~ nutrient_level + species,
                         data = algae, FUN = mean)
```

```{r event-setup, context="server-start"}
event_register_handler("exercise_result", function(session, event, data) {
  log_attempt(
    student_id  = bx_get_sid(session),
    exercise_id = data$label,
    correct     = data$feedback$correct
  )
})
```

```{r coach-server, context="server"}
# One coach_server() per coached exercise. The third argument must match a
# key in the server-side exercise_registry.R ("moduleN-qM", HYPHENATED) --
# an unknown key does not error, it silently returns generic context.
# The Shiny id must match the coach_ui() call below it, or the card renders
# and never responds.
BisonExplorR::coach_server("coach_q1", shiny::reactive(learnr::get_tutorial_state("q1")), "module7-q1")
BisonExplorR::coach_server("coach_q3", shiny::reactive(learnr::get_tutorial_state("q3")), "module7-q3")
```


## Before you start

```{r student-id, echo=FALSE}
uiOutput("sid_ui")
```

```{r sid-server, context="server"}
# The ID box exists ONLY in the graded deployment. See bx_is_graded() in setup.
output$sid_ui <- renderUI({
  if (!bx_is_graded()) {
    return(shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #C05621; background:#FDF6EC;"),
      shiny::strong("Practice mode."),
      " Nothing here is recorded and this copy does not count for credit.",
      " To do this module for credit, run ",
      shiny::tags$code('launch_graded_tutorial("module7")'),
      " instead."
    ))
  }
  shiny::tagList(
    shiny::div(
      style = paste("padding:12px 14px; margin:6px 0;",
                    "border-left:4px solid #1F3864; background:#F2F4F7;"),
      shiny::strong("Graded copy."),
      " Your work on this module is recorded, so enter your student ID before",
      " you begin. Use the same ID every time \u2014 a different ID reads as a",
      " different student and your work will not be combined."
    ),
    shiny::textInput("sid",
      "Student ID (the same one you used in Module 1):", value = ""),
    shiny::div(id = "sid_warn",
      style = "color:#C05621; font-weight:600; margin:-6px 0 8px 0;",
      "Enter your ID to continue."),
    # Client-side gate on the section's Continue button. This prevents the
    # ACCIDENTAL case (a student who forgets), not a determined one -- anyone
    # with devtools can re-enable the button. The real protection against
    # unattributed rows is that grade_module() drops blank IDs.
    shiny::tags$script(shiny::HTML("
      (function () {
        function apply() {
          var sid = document.getElementById('sid');
          if (!sid) return;
          var sec = sid.closest('.section');
          if (!sec) return;
          var btns = sec.querySelectorAll('.topicActions button, .continueButton');
          var ok = sid.value.trim().length > 0;
          btns.forEach(function (b) {
            b.disabled = !ok;
            b.style.opacity = ok ? '' : '0.45';
            b.style.cursor  = ok ? '' : 'not-allowed';
          });
          var w = document.getElementById('sid_warn');
          if (w) w.style.display = ok ? 'none' : 'block';
        }
        document.addEventListener('input', function (e) {
          if (e.target && e.target.id === 'sid') apply();
        });
        // learnr renders sections progressively, so the Continue button may not
        // exist yet when this script first runs.
        new MutationObserver(apply)
          .observe(document.body, { childList: true, subtree: true });
        apply();
      })();
    "))
  )
})

# Cache the ID on userData as a second, independent path to it. bx_get_sid()
# checks rootScope() first and falls back to this.
shiny::observeEvent(input$sid, {
  if (!is.null(input$sid) && nzchar(input$sid)) {
    session$userData$bx_sid <- input$sid
  }
}, ignoreInit = TRUE)
```


## The map: which test, when

Keep this table in view — it is the same in every statistics module. The last
column tells you where each test is taught.

| Biological question | Predictor(s) | Test | R function | Module |
|---|---|---|---|---|
| Do **two** groups differ? | one factor, 2 levels | t-test | `t.test()` | 5 |
| Do **three or more** groups differ? | one factor, 3+ levels | one-way ANOVA | `aov()` + `summary()` | 5 |
| Does a **continuous** variable predict the response? | one continuous | linear regression | `lm()` + `summary()` | 6 |
| Do groups differ across **two factors**? | two factors | two-way ANOVA | `aov()` + `summary()` | 7 |
| Do groups differ **after controlling for** a continuous variable? | factor + covariate | ANCOVA | `aov()` + `summary()` | 8 |

**Where you are:** Module 7 — two grouping factors at once.

## One response, two grouping factors

You have seen this data before. In **Module 5** these were the same 60 vials of
alginate-encapsulated algae beads: two species, three nutrient levels, and CO2
drawdown measured for each vial.

But in Module 5 you asked two *separate* questions:

- a **t-test** — does drawdown differ between the two species?
- a **one-way ANOVA** — does drawdown differ across the three nutrient levels?

Each of those threw away half the design. The t-test pooled all three nutrient
levels together and called that variation noise; the ANOVA did the same to
species. A **two-way ANOVA** fits both factors in one model instead.

It is still `aov()` + `summary()` — just two predictors instead of one. Writing
the factors with a `*` between them does something extra: it also estimates
their **interaction**, whether the effect of one factor *depends on* the level
of the other. `nutrient_level * species` is shorthand for both main effects plus
the interaction `nutrient_level:species`.

```{r peek, exercise=TRUE}
head(algae)
```

## Fit the two-way ANOVA {data-allow-skip="true"}

### Exercise q1

Fit a two-way ANOVA of `co2_drawdown_ppm` by `nutrient_level` and `species`
**with their interaction**, and print the summary.

```{r q1, exercise=TRUE}
summary(aov(co2_drawdown_ppm ~ ___ * ___, data = algae))
```

```{r q1-hint}
# The * gives you both main effects and the interaction.
summary(aov(co2_drawdown_ppm ~ nutrient_level * species, data = algae))
```

```{r q1-solution}
summary(aov(co2_drawdown_ppm ~ nutrient_level * species, data = algae))
```

```{r q1-check}
grade_this({
  if (inherits(.result, "summary.aov")) {
    terms <- rownames(.result[[1]])
    if (any(grepl("nutrient", terms)) && any(grepl("species", terms)) && any(grepl(":", terms))) {
      pass("Three rows above Residuals: two main effects and the interaction. Nutrient F = 26.75, species F = 10.31, interaction F = 0.13. Always read the interaction line first.")
    }
    if (any(grepl("nutrient", terms)) && any(grepl("species", terms))) {
      fail("Close — both factors are in the model, but there is no interaction row. Use * between them, not +.")
    }
  }
  fail("Not quite. Use summary(aov(co2_drawdown_ppm ~ nutrient_level * species, data = algae)) — the * adds the interaction.")
})
```

```{r coach-ui-q1, echo=FALSE}
BisonExplorR::coach_ui("coach_q1", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

### Read the interaction first

The interaction row tells you whether the main effects can even be interpreted
on their own, so it comes first. Here `nutrient_level:species` has
**p = 0.881** — nowhere near significant — while both main effects are
significant (nutrient p = 8.4e-09, species p = 0.0022).

```{r q1_interp, echo=FALSE}
question(
  "The interaction has p = 0.881 and both main effects are significant. What does that license you to say?",
  answer("The effect of nutrient level is the same for both species, so the two main effects can be interpreted on their own.", correct = TRUE),
  answer("Neither nutrient level nor species matters.", message = "Both main effects are significant — p = 8.4e-09 and p = 0.0022. It is only the interaction that is not."),
  answer("You must drop a main effect from the model.", message = "A non-significant interaction is a result, not a reason to remove a main effect. Both factors stay."),
  answer("Nutrient level and species are measuring the same thing.", message = "They are separate factors. The interaction asks whether their effects combine, not whether they are identical."),
  allow_retry = TRUE
)
```

### Why the two-way is worth the trouble

Compare what you just fitted to Module 5:

| Question | Module 5 (one factor) | Module 7 (both factors) |
|---|---|---|
| Does species matter? | t-test, **p = 0.022** | **p = 0.0022** |
| Does nutrient matter? | one-way ANOVA, **p = 3.4e-08** | **p = 8.4e-09** |

Same data, same 60 vials, and **both effects came out stronger.** Nothing was
added — something was *removed*. In Module 5, when you tested species alone, all
the variation caused by nutrient level was left sitting in the error term, making
the noise look bigger than it really was. Fitting both factors together explains
that variation instead of dumping it in the residual, so each effect is measured
against a cleaner baseline.

That is the practical argument for a two-way design: it is not just two tests at
once, it is two **more sensitive** tests.

## Show both factors at once {data-allow-skip="true"}

A two-way design needs a figure that shows *both* factors. Two work well: a
**grouped boxplot** (raw distributions, split by both factors) and an
**interaction plot** (group means connected by lines).

### Exercise q2: grouped boxplot

Build a boxplot of `co2_drawdown_ppm` by `nutrient_level`, with the boxes split
by `species` using `fill`. Finish with `theme_classic()`.

```{r q2, exercise=TRUE}
ggplot(algae, aes(x = nutrient_level, y = co2_drawdown_ppm, fill = ___)) +
  geom_boxplot() +
  labs(x = "Nutrient level", y = "CO2 drawdown (ppm)") +
  theme_classic()
```

```{r q2-hint}
# Map fill to the second factor.
# fill = species
```

```{r q2-solution}
ggplot(algae, aes(x = nutrient_level, y = co2_drawdown_ppm, fill = species)) +
  geom_boxplot() +
  labs(x = "Nutrient level", y = "CO2 drawdown (ppm)") +
  theme_classic()
```

```{r q2-check}
grade_this({
  if (inherits(.result, "ggplot")) {
    aes_names <- names(.result$mapping)
    geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
    if (any(geoms == "GeomBoxplot") && "fill" %in% aes_names) {
      pass("Now every nutrient level shows both species side by side. Both main effects are visible in one panel: the high boxes sit above the rest, and within every level the Scenedesmus box sits above the Chlorella box.")
    }
    if (any(geoms == "GeomBoxplot")) {
      fail("You have the boxplot, but only one factor is showing. Map fill = species inside aes() so both factors appear.")
    }
  }
  fail("Not quite. Map fill = species inside aes() and use geom_boxplot().")
})
```

### Exercise q3: interaction plot

An interaction plot draws the **group means** and connects them with a line per
group; nearly parallel lines signal little interaction. The means are already
computed for you in `algae_means`. Map `colour = species` and `group = species`,
then add `geom_point()` and `geom_line()`.

```{r q3, exercise=TRUE}
ggplot(algae_means, aes(x = nutrient_level, y = co2_drawdown_ppm, colour = species, group = ___)) +
  geom_point() +
  geom_line() +
  labs(x = "Nutrient level", y = "Mean CO2 drawdown (ppm)") +
  theme_classic()
```

```{r q3-hint}
# group = species tells ggplot to connect one line per species.
# Without it ggplot sees a discrete x axis, finds no grouping, and draws nothing.
```

```{r q3-solution}
ggplot(algae_means, aes(x = nutrient_level, y = co2_drawdown_ppm, colour = species, group = species)) +
  geom_point() +
  geom_line() +
  labs(x = "Nutrient level", y = "Mean CO2 drawdown (ppm)") +
  theme_classic()
```

```{r q3-check}
grade_this({
  if (inherits(.result, "ggplot")) {
    geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
    if (any(geoms == "GeomLine") && any(geoms == "GeomPoint")) {
      pass("The two lines climb together and never cross — the visual signature of the non-significant interaction you found in q1. Scenedesmus draws down more at every nutrient level, and both species respond to nutrients the same way.")
    }
    if (any(geoms == "GeomPoint")) {
      fail("The points are there but the lines are missing. Check that group = species is set — geom_line() needs to know which points to connect.")
    }
  }
  fail("Not quite. Set group = species, then add geom_point() and geom_line().")
})
```

```{r coach-ui-q3, echo=FALSE}
BisonExplorR::coach_ui("coach_q3", "Stuck? Ask the Coach")
```

> The Coach reads the attempt you last **submitted** — it will not see code you only ran. It gives one nudge, never the answer.

### What a significant interaction would have looked like

Your lines run roughly parallel. If the interaction *had* been significant, they
would converge, diverge, or cross — meaning "does nutrient level matter?" would
have no single answer, because the answer would be *it depends on the species*.
That is why the interaction is read first: a significant one makes the main
effect rows misleading, not just incomplete.

## Where you landed

- **Two grouping factors →** `summary(aov(response ~ factorA * factorB))`
- The `*` adds the **interaction**; read that line **first**
- A non-significant interaction means the main effects stand on their own
- Fitting both factors together is **more sensitive** than two separate tests
- Show both factors with a **grouped boxplot** or an **interaction plot**

Next module keeps two predictors but changes the type of the second one: instead
of a grouping factor, a **continuous** measurement you want to control for. That
is **ANCOVA**, and it combines the regression you did in Module 6 with the group
comparison you did in Module 5.
